Stability of variational quantum dynamics with Gaussian approximants

### 2013-07-11 paper.tex
writing

When the first quantum physicists realised that the position of a particle is essentially uncertain, they were destined to seek the conditions that cause it to follow a normal law of error, and discover the gaussian wave packet.  They soon understood the free particle and turned to the harmonic oscillator, investigating its ground state and how its excited states generate sinusoidal motion in the classical limit.  These problems were also solved by the gaussian wave packet.  With its width fixed by equipartition of energy and the uncertainty principle, it became the coherent state.

The simplest mechanical systems allow Schrödinger's law of motion to be solved in closed form, but this luck doesn't last.  Most quantum problems must be solved approximately.  This can be done variationally, specifying a class of approximants that can be manipulated exactly, then finding one of them that most nearly solves the problem.  To test how well a putative solution approximates the dynamics of a system, it can be substituted into each side of Schrödinger's equation, and the difference between them evaluated by the usual norm in Hilbert space.  It remains to choose a class of approximants, and the gaussian wave packets are an common choice.  They differ in their scale and their central position and momentum, useful properties to determine for any quantum state.

The laser was invented, and physicists needed a quantum theory of intense radiation.  This obeyed simple laws of dynamics, very similar to harmonic oscillators.  However, It involved up to $10^{18}$ photons, spread across thousands of modes, so the difficulty of the quantum problem came from its sheer size.  This was made managable by the work of Glauber, who showed that any quantum state could be written as a superposition of coherent states.  For optical systems, close to harmonic oscillators, this was a very efficient way to represent states.  Moyal(?) developed an efficient way to calculate dynamics, leading to a family of phase space Monte-Carlo methods.  These have solved many problems, but are now limited by intractable convergence problems.

In the 1970s, Heller put the two ideas together, and proposed simulating quantum dynamics variationally, using superpositions of coherent states as the approximant.  This is an appealing approach.  Every quantum state can be approximated arbitrarily well as the number of components in the superposition becomes large.  The experience of quantum optics shows that many systems can be approximated by a reasonable number of components, and variational dynamics should not have the covergence problems the Monte-Carlo methods do.

@@

Since Heller suggested it, there have been several attempts to apply this numerically.  In all cases, numerical instability has prevented the superposition being expanded to the point of convergence.
%% Consequence
These attempts have aimed to solve chemistry problems, involving dozens of atoms, where the lack of convergence limited the precision that could be obtained, but superpositions with a few components, and ad-hoc methods to improve the stability, still gave useful results.  The kind of problems that people approach with phase space methods come from atom optics or condensed matter; they involve thousands or millions of atoms, and the convergence problems would prevent any useful information being obtained.  A systematic approach to regularisation is required in order to approach these problems variationally.  The prospective rewards are great: these problems include high temperature superconductivity and the dynamics of the early universe, topics on which today's physicists expend massive efforts.

>>>>

When the first quantum physicists realised that the position of a particle is essentially uncertain, they were destined to seek the conditions that cause it to follow a normal law of error, and discover the gaussian wave packet.  They soon understood the free particle and turned to the harmonic oscillator, investigating the form of its ground state and the way its excited states generate sinusoidal motion in the classical limit.  These problems were also solved by the gaussian wave packet.  With its width fixed by equipartition of energy and the uncertainty principle, it became the coherent state.

The simplest mechanical systems allow Schrödinger's law of motion to be solved in closed form, but most quantum problems must be solved approximately.  This can be done variationally, specifying a class of approximants that can be manipulated exactly, then finding one of them that most nearly solves the problem.  To test how well a wave function approximates the dynamics of a system, the difference between the two sides of Schrödinger's equation can be measured by the Hilbert space norm.  It remains to choose a class of approximants.  The gaussian wave packets are an common choice, because they differ in their scale and their central position and momentum, useful properties to determine for any quantum state.

The invention of the laser called for a quantum theory of intense radiation.  Laser beams are difficult because their quantum descriptions are so large: they involve up to $10^{18}$ photons, distributed across thousands of modes.  The theory was made managable by Glauber and Sudarshan, who showed that any quantum state can be expanded as a superposition of coherent states.  Laser light is very coherent, so these expansions often represent it efficiently.  Moyal(?) developed an efficient way to calculate dynamics, leading to a family of phase space Monte-Carlo methods.  When atomic Bose-Einstein condensates were made, and very coherent matter became easy to access in the lab, phase space methods were sucessfully applied to calculate its dynamics, and have since been applied to other forms of condensed matter.  They have solved many problems, but are limited by intractable convergence issues.

In the 1970s, Heller put the two ideas together, proposing to simulate quantum dynamics variationally, using finite superpositions of coherent states as the approximants.  This is an appealing approach.  Every quantum state can be approximated arbitrarily well as the number of components in the superposition increases.  The experience of quantum optics shows that many systems can be approximated by a reasonable number of components, and variational dynamics should avoid the covergence problems that limit the Monte-Carlo methods.

@@

Since Heller suggested it, there have been several attempts to apply this numerically, but none has entirely succeeded.  In all cases, numerical instability has prevented the superposition being expanded to the point of convergence.
%% Consequence
These attempts have aimed to solve chemistry problems, involving dozens of atoms, where the lack of convergence limited the precision that could be obtained, but superpositions with a few components, and ad-hoc methods to improve the stability, still gave useful results.  The kind of problems that people approach with phase space methods come from atom optics or condensed matter; they involve thousands or millions of atoms, and the convergence problems would prevent any useful information being obtained.  A systematic approach to regularisation is required in order to approach these problems variationally.  The prospective rewards are great: these problems include high temperature superconductivity and the dynamics of the early universe, topics on which today's physicists expend massive efforts.
